---
title: "extracting the isiccomb -> isic crossmap collection from INDSTAT"
execute:
  warning: false
  cache: true
---

## Motivation

This page is a companion to
[`xmap`](https://github.com/cynthiahqy/xmap)'s vignette
`extracting-crossmaps-from-scripts.Rmd` (specifically Case 2, PR #4) --
it runs the same "carbon paper" naive-weight extraction technique end to
end on the **full, real INDSTAT Rev.3 extract** (all reporters and years,
not the vignette's 5-country masked-sample subset), and saves the result
as a single artifact the rest of this site reads from.

The technique, restated: rather than trusting or parsing a legacy
data-cleaning script's logic directly, pass "simplified" (masked) data
through it and recover the mapping it implies from the output. This works
even when the real values are sensitive, since only the *structure* of
the transform -- which source codes map to which targets, at what weight
-- is being recovered, not the real data.

**Every other page in this site loads `data/interim/crossmap_collection.rds`,
produced at the end of this page, rather than recomputing any of this.**
There is no build-system dependency graph (no `targets`/`_targets.R`) --
this is deliberately a single, linear, readable extraction notebook, in
keeping with this site's purpose as a companion to the vignette and to
the crossmaps definition paper, not a general-purpose data pipeline. If
you change this page's logic, re-render it before the other pages, or
just `quarto render` the whole project (files are numbered so this one
sorts and renders first).

## Setup

```{r}
library(tidyverse)
library(here)
library(xmap)
library(testthat)
```

## The legacy script

Treated as an opaque black box, exactly as in the vignette -- we recover
its mapping logic by extraction rather than by reading it:

```{r}
#| file: R/legacy.R
#| eval: true
```

## Read the raw INDSTAT extract

```{r}
#| file: R/load.R
#| eval: true
```

```{r}
ctr_codes <- read_rev3_ctr_codes(here("data", "external", "country_codes", "INDSTAT_countries_REV3.csv"))
indstat_rev3 <- read_rev3_output(here("data", "raw", "REV_3", "14-Output.csv"))
indstat_threefour <- subset_34_from_total(indstat_rev3, total_letter = "D")$threefour

indstat_threefour
```

## Extract the naive weights

Masking must preserve `NA`s -- an unconditional mask would turn
originally-missing cells into "recorded" ones, breaking
`split_isiccomb()`'s own internal invariant check (exactly one recorded
value per `isiccomb` group). See the comment in `R/extract.R`; this was a
real bug caught while building this page, not a theoretical concern.

```{r}
#| file: R/extract.R
#| eval: true
```

```{r}
links_isiccomb_isic <- extract_naive_weights(indstat_threefour)

links_isiccomb_isic
```

## Validate, per country-year

A crossmap collection is many small crossmaps stacked in one table, so
validity has to be checked *within* each `(country, year)` group, not
across the whole table at once:

```{r}
diagnoses_isiccomb_isic <- diagnose_grouped_xmap(links_isiccomb_isic, isiccomb, isic, weight)

diagnoses_isiccomb_isic |> count(valid)
```

## Derive the collection's summary artifacts

The deterministic hierarchical map `isic -> isic.3` (weight always `1`;
kept separate from the split map above so a future
`xmap::compose_xmap()` -- [xmap#29](https://github.com/cynthiahqy/xmap/issues/29)
-- can chain the two without this page needing to change):

```{r}
links_isic_isic3 <- links_isiccomb_isic |>
  distinct(isic) |>
  mutate(isic3 = str_sub(isic, 1, 3), weight = 1)
```

**Stage-1 coverage tiles** -- these only look at the split
(`isiccomb -> isic`) in isolation: does a country-year have any split at
all, and what fraction of `isic` values were touched by one. Neither
knows `isic.3` exists yet:

```{r}
tile_any_split <- links_isiccomb_isic |>
  group_by(country, year) |>
  summarise(any_split = any(weight != 1), .groups = "drop")

tile_frac_split <- links_isiccomb_isic |>
  group_by(country, year) |>
  summarise(
    n_isic = n(),
    n_split = sum(weight != 1),
    frac_split = n_split / n_isic,
    .groups = "drop"
  )
```

Structural `(country, year, isiccomb, isic, isic3, weight)` edge table
for the node-link diagrams:

```{r}
combo_edges <- links_isiccomb_isic |>
  filter(split.isiccomb == TRUE) |>
  transmute(country, year, isiccomb, isic, isic3 = str_sub(isic, 1, 3), weight)

combo_edges
```

Which `(country, year, isiccomb)` groups cross an `isic.3` boundary
(span more than one distinct `isic.3` parent):

```{r}
comb_group_span <- combo_edges |>
  group_by(country, year, isiccomb) |>
  summarise(n_isic3 = n_distinct(isic3), .groups = "drop") |>
  mutate(crosses_isic3 = n_isic3 > 1)
```

Is boundary-crossing a property of the `isiccomb` *code* itself (a fixed
classification artefact, true wherever the code is reported) or of the
reporting context (varies by which country/year uses it)? Pooling across
*all* country-years and looking at each code's membership structurally
answers this directly:

```{r}
comb_code_span <- combo_edges |>
  distinct(isiccomb, isic, isic3) |>
  group_by(isiccomb) |>
  summarise(
    n_isic3 = n_distinct(isic3),
    isic3_members = paste(sort(unique(isic3)), collapse = ", "),
    .groups = "drop"
  ) |>
  mutate(crosses_isic3 = n_isic3 > 1) |>
  arrange(desc(n_isic3))

comb_code_span
```

It's overwhelmingly a code property, not a reporting-context one: of
306 distinct `isiccomb` codes, 179 (58%) always cross wherever they're
reported, and boundary-crossing components make up 86% of all
`(country, year, isiccomb)` groups. Some codes are extreme --
`281I` spans 25 different `isic.3` parents every time it appears:

```{r}
comb_code_span |>
  summarise(n_codes = n(), n_codes_crossing = sum(crosses_isic3)) |>
  mutate(pct_codes_crossing = 100 * n_codes_crossing / n_codes)

comb_code_span |> filter(isiccomb == "281I")
```

Composing the split with the deterministic `isic -> isic.3` aggregation
can do one of two things to a combo code's children: **reconverge**
(every child lands back in the same `isic.3` parent -- the `1/n` split
weights cancel out exactly on re-aggregation, so the `isic.3`-level
figure is mathematically exact regardless of the split), or **cross**
(children land in *different* `isic.3` parents -- now the even-split
assumption directly determines how much value each parent receives, with
no reconvergence to cancel it out). Only crossing components carry real,
propagated allocation uncertainty past the split step; reconverging ones
are imputed at the 4-digit level but exact at 3-digit and coarser. This
is exactly what the node-link diagrams in
[Explorer](02-crossmap-explorer.qmd)/[Country variants](03-country-variants.qmd)
show per component (dashed split edges into a reconvergent or diverging
fan); `crosses_isic3` above is that same distinction at the
per-`(country, year, isiccomb)` grain.

**Stage-1+2 coverage tile**: `frac_split_crossing` is the same
denominator as `frac_split` above (fraction of a country-year's `isic`
values), restricted to values whose combo group crosses an `isic.3`
boundary -- i.e. a decomposition, `frac_split = frac_split_reconverging +
frac_split_crossing`, isolating the fraction of "touched by imputation"
that's actually consequential once you account for the second stage:

```{r}
crossing_isic_rows <- combo_edges |>
  distinct(country, year, isiccomb, isic) |>
  inner_join(
    comb_group_span |> filter(crosses_isic3) |> select(country, year, isiccomb),
    by = c("country", "year", "isiccomb")
  )

tile_frac_split_crossing <- tile_frac_split |>
  select(country, year, n_isic) |>
  left_join(
    crossing_isic_rows |> count(country, year, name = "n_crossing"),
    by = c("country", "year")
  ) |>
  mutate(
    n_crossing = replace_na(n_crossing, 0),
    frac_split_crossing = n_crossing / n_isic
  )

tile_frac_split_crossing
```

Exact per-`(country, year)` crossmap signature -- sorted, concatenated
`isiccomb|isic|weight` triples, so identical strings mean identical
crossmaps -- plus summary stats, used by both explorer pages to
de-duplicate before visualizing:

```{r}
#| file: R/prep.R
#| eval: true
```

```{r}
ctr_year_signature <- compute_ctr_year_signature(combo_edges, comb_group_span, ctr_codes)

ctr_year_signature |> select(-signature)
```

Global max `isic.3` in-degree (max # of distinct `isic` children feeding
into a single `isic.3` parent, over *all* country-year-`isiccomb`
components in the dataset) -- a fixed reference point for the node-link
diagrams' target-node color scale, so "how dark" means the same thing in
every diagram on the site rather than being renormalized per diagram:

```{r}
max_isic3_indegree <- combo_edges |>
  group_by(country, year, isiccomb, isic3) |>
  summarise(n_isic = n_distinct(isic), .groups = "drop") |>
  pull(n_isic) |>
  max()

max_isic3_indegree
```

## Save the collection

```{r}
dir.create(here("data", "interim"), showWarnings = FALSE, recursive = TRUE)

crossmap_collection <- list(
  ctr_codes = ctr_codes,
  links_isiccomb_isic = links_isiccomb_isic,
  links_isic_isic3 = links_isic_isic3,
  diagnoses_isiccomb_isic = diagnoses_isiccomb_isic,
  tile_any_split = tile_any_split,
  tile_frac_split = tile_frac_split,
  tile_frac_split_crossing = tile_frac_split_crossing,
  combo_edges = combo_edges,
  comb_group_span = comb_group_span,
  comb_code_span = comb_code_span,
  ctr_year_signature = ctr_year_signature,
  max_isic3_indegree = max_isic3_indegree
)

saveRDS(crossmap_collection, here("data", "interim", "crossmap_collection.rds"))
```
