---
title: "extracting the isiccomb -> isic crossmap collection from INDSTAT"
execute:
  warning: false
  cache: true
---

## Motivation

This page is a companion to
[`xmap`](https://github.com/cynthiahqy/xmap)'s vignette
`extracting-crossmaps-from-scripts.Rmd` (specifically Case 2, PR #4) --
it runs the same "carbon paper" naive-weight extraction technique end to
end on the **full, real INDSTAT Rev.3 extract** (all reporters and years,
not the vignette's 5-country masked-sample subset), and saves the result
as a single artifact the rest of this site reads from.

The technique, restated: rather than trusting or parsing a legacy
data-cleaning script's logic directly, pass "simplified" (masked) data
through it and recover the mapping it implies from the output. This works
even when the real values are sensitive, since only the *structure* of
the transform -- which source codes map to which targets, at what weight
-- is being recovered, not the real data.

**Every other page in this site loads `data/interim/crossmap_collection.rds`,
produced at the end of this page, rather than recomputing any of this.**
There is no build-system dependency graph (no `targets`/`_targets.R`) --
this is deliberately a single, linear, readable extraction notebook, in
keeping with this site's purpose as a companion to the vignette and to
the crossmaps definition paper, not a general-purpose data pipeline. If
you change this page's logic, re-render it before the other pages, or
just `quarto render` the whole project (files are numbered so this one
sorts and renders first).

## Setup

```{r}
library(tidyverse)
library(here)
library(xmap)
library(testthat)
```

## The legacy script

Treated as an opaque black box, exactly as in the vignette -- we recover
its mapping logic by extraction rather than by reading it:

```{r}
#| file: R/legacy.R
#| eval: true
```

## Read the raw INDSTAT extract

```{r}
#| file: R/load.R
#| eval: true
```

```{r}
ctr_codes <- read_rev3_ctr_codes(here("data", "external", "country_codes", "INDSTAT_countries_REV3.csv"))
indstat_rev3 <- read_rev3_output(here("data", "raw", "REV_3", "14-Output.csv"))
indstat_threefour <- subset_34_from_total(indstat_rev3, total_letter = "D")$threefour

indstat_threefour
```

## Extract the naive weights

Masking must preserve `NA`s -- an unconditional mask would turn
originally-missing cells into "recorded" ones, breaking
`split_isiccomb()`'s own internal invariant check (exactly one recorded
value per `isiccomb` group). See the comment in `R/extract.R`; this was a
real bug caught while building this page, not a theoretical concern.

```{r}
#| file: R/extract.R
#| eval: true
```

```{r}
links_isiccomb_isic <- extract_naive_weights(indstat_threefour)

links_isiccomb_isic
```

## Validate, per country-year

A crossmap collection is many small crossmaps stacked in one table, so
validity has to be checked *within* each `(country, year)` group, not
across the whole table at once:

```{r}
diagnoses_isiccomb_isic <- diagnose_grouped_xmap(links_isiccomb_isic, isiccomb, isic, weight)

diagnoses_isiccomb_isic |> count(valid)
```

## Derive the collection's summary artifacts

The deterministic hierarchical map `isic -> isic.3` (weight always `1`;
kept separate from the split map above so a future
`xmap::compose_xmap()` -- [xmap#29](https://github.com/cynthiahqy/xmap/issues/29)
-- can chain the two without this page needing to change):

```{r}
links_isic_isic3 <- links_isiccomb_isic |>
  distinct(isic) |>
  mutate(isic3 = str_sub(isic, 1, 3), weight = 1)
```

Coverage tiles -- does a country-year have any split at all, and what
fraction of `isic` values were touched by one:

```{r}
tile_any_split <- links_isiccomb_isic |>
  group_by(country, year) |>
  summarise(any_split = any(weight != 1), .groups = "drop")

tile_frac_split <- links_isiccomb_isic |>
  group_by(country, year) |>
  summarise(
    n_isic = n(),
    n_split = sum(weight != 1),
    frac_split = n_split / n_isic,
    .groups = "drop"
  )
```

Structural `(country, year, isiccomb, isic, isic3, weight)` edge table
for the node-link diagrams:

```{r}
combo_edges <- links_isiccomb_isic |>
  filter(split.isiccomb == TRUE) |>
  transmute(country, year, isiccomb, isic, isic3 = str_sub(isic, 1, 3), weight)

combo_edges
```

Which `(country, year, isiccomb)` groups cross an `isic.3` boundary
(span more than one distinct `isic.3` parent):

```{r}
comb_group_span <- combo_edges |>
  group_by(country, year, isiccomb) |>
  summarise(n_isic3 = n_distinct(isic3), .groups = "drop") |>
  mutate(crosses_isic3 = n_isic3 > 1)
```

Exact per-`(country, year)` crossmap signature -- sorted, concatenated
`isiccomb|isic|weight` triples, so identical strings mean identical
crossmaps -- plus summary stats, used by both explorer pages to
de-duplicate before visualizing:

```{r}
#| file: R/prep.R
#| eval: true
```

```{r}
ctr_year_signature <- compute_ctr_year_signature(combo_edges, comb_group_span, ctr_codes)

ctr_year_signature |> select(-signature)
```

Global max `isic.3` in-degree (max # of distinct `isic` children feeding
into a single `isic.3` parent, over *all* country-year-`isiccomb`
components in the dataset) -- a fixed reference point for the node-link
diagrams' target-node color scale, so "how dark" means the same thing in
every diagram on the site rather than being renormalized per diagram:

```{r}
max_isic3_indegree <- combo_edges |>
  group_by(country, year, isiccomb, isic3) |>
  summarise(n_isic = n_distinct(isic), .groups = "drop") |>
  pull(n_isic) |>
  max()

max_isic3_indegree
```

## Save the collection

```{r}
dir.create(here("data", "interim"), showWarnings = FALSE, recursive = TRUE)

crossmap_collection <- list(
  ctr_codes = ctr_codes,
  links_isiccomb_isic = links_isiccomb_isic,
  links_isic_isic3 = links_isic_isic3,
  diagnoses_isiccomb_isic = diagnoses_isiccomb_isic,
  tile_any_split = tile_any_split,
  tile_frac_split = tile_frac_split,
  combo_edges = combo_edges,
  comb_group_span = comb_group_span,
  ctr_year_signature = ctr_year_signature,
  max_isic3_indegree = max_isic3_indegree
)

saveRDS(crossmap_collection, here("data", "interim", "crossmap_collection.rds"))
```
