---
title: "coverage: which country-years rely on a split?"
execute:
  warning: false
  cache: true
---

## Motivation

Two different questions, at two different stages of the transformation:

- **Stage 1** (below): was this `isic` value touched by an `isiccomb`
  split *at all*? This only looks at `isiccomb -> isic` in isolation --
  it has no notion of `isic.3` and can't distinguish a split that's
  harmless from one that isn't.
- **Stage 1+2** (further below): composing the split with the
  deterministic `isic -> isic.3` aggregation, does the combined
  transformation actually cross an `isic.3` boundary? A split whose
  children all reconverge into the same `isic.3` parent is imputed at
  the 4-digit level but **exact** at 3-digit and coarser (the `1/n`
  weights cancel out on re-aggregation); a split whose children land in
  *different* `isic.3` parents carries real, propagated allocation
  uncertainty. Only the second kind should really count as "consequential"
  imputation exposure. This is the same distinction the node-link
  diagrams in [Explorer](02-crossmap-explorer.qmd)/
  [Country variants](03-country-variants.qmd) show per component (a
  reconvergent vs. diverging fan of dashed split edges); this page shows
  its coverage-level rollup.

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
tile_any_split <- crossmap_collection$tile_any_split
tile_frac_split <- crossmap_collection$tile_frac_split
tile_frac_split_crossing <- crossmap_collection$tile_frac_split_crossing
ctr_codes <- crossmap_collection$ctr_codes
```

All three tiles below share one country ordering -- by each country's
*overall* (all-years-pooled) fraction of `isic` values in a crossing
component, weighted by volume (`sum(n_crossing) / sum(n_isic)`, not a
plain per-year average, so a country's one thin reporting year can't
skew its rank) -- rather than each plot re-sorting on its own metric.
That keeps a given row meaning "this country" consistently across all
three figures, so you can read the same row down the page and see
stage-1 coverage vs. stage-1+2 consequential coverage for the same
country side by side.

```{r}
country_order <- tile_frac_split_crossing |>
  group_by(country) |>
  summarise(total_isic = sum(n_isic), total_crossing = sum(n_crossing), .groups = "drop") |>
  mutate(agg_frac_crossing = total_crossing / total_isic) |>
  arrange(agg_frac_crossing) |>
  pull(country)
```

## Stage 1: split coverage (isiccomb -> isic only)

Two views of the same underlying fact, at different resolutions: the
binary tile is a strict coarsening of the continuous one
(`any_split == (frac_split > 0)`) -- read them as a pair, binary for
"where does this even come up," continuous for "how much does it matter
there."

### Binary: does this country-year have any isiccomb split at all?

```{r}
tile_any_split |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  mutate(country = factor(country, levels = country_order)) |>
  ggplot(aes(x = as.factor(year), y = country, fill = any_split)) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_manual(values = c("TRUE" = "#ef8a62", "FALSE" = "#67a9cf")) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_blank()) +
  labs(
    x = "Year", y = "Country", fill = "Any split?",
    title = "Countries reporting isiccomb combination codes",
    subtitle = "INDSTAT 4 REV. 3 Output data -- stage 1 only"
  )
```

### Continuous: what fraction of reported isic values came from a split?

```{r}
tile_frac_split |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  mutate(country = factor(country, levels = country_order)) |>
  ggplot(aes(x = as.factor(year), y = country, fill = frac_split)) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_blank()) +
  labs(
    x = "Year", y = "Country", fill = "Fraction split",
    title = "Share of isic values derived from an isiccomb split",
    subtitle = "INDSTAT 4 REV. 3 Output data -- stage 1 only"
  )
```

## Stage 1+2: does the transformation actually cross an isic.3 boundary?

`frac_split_crossing` shares `frac_split`'s denominator (fraction of a
country-year's `isic` values) but restricts the numerator to values
whose combo group *crosses* an `isic.3` boundary -- i.e. it's the
"consequential" subset of stage 1's coverage, once the second stage is
accounted for: `frac_split = frac_split_reconverging + frac_split_crossing`.

```{r}
tile_frac_split_crossing |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  mutate(country = factor(country, levels = country_order)) |>
  ggplot(aes(x = as.factor(year), y = country, fill = frac_split_crossing)) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_distiller(palette = "Oranges", direction = 1) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_blank()) +
  labs(
    x = "Year", y = "Country", fill = "Fraction crossing",
    title = "Share of isic values with real, propagated allocation uncertainty",
    subtitle = "INDSTAT 4 REV. 3 Output data -- stage 1+2 composed"
  )
```

How much of stage 1's coverage is actually consequential, in aggregate
across the whole dataset:

```{r}
overall <- tile_frac_split |>
  left_join(tile_frac_split_crossing |> select(country, year, frac_split_crossing),
    by = c("country", "year")
  ) |>
  summarise(
    total_isic = sum(n_isic),
    total_split = sum(n_split),
    total_crossing = sum(round(frac_split_crossing * n_isic)),
  )

overall |>
  mutate(
    pct_of_all_isic_split = 100 * total_split / total_isic,
    pct_of_all_isic_crossing = 100 * total_crossing / total_isic,
    pct_of_split_that_crosses = 100 * total_crossing / total_split
  )
```
