---
title: "Split Coverage"
execute:
  warning: false
  cache: true
---

## Motivation

Two different questions, at two different stages of the transformation:

- **Stage 1** (below): was this `isic` value touched by an `isiccomb`
  split *at all*? This only looks at `isiccomb -> isic` in isolation --
  it has no notion of `isic.3` and can't distinguish a split that's
  harmless from one that isn't.
- **Stage 1+2** (further below): composing the split with the
  deterministic `isic -> isic.3` aggregation, does the combined
  transformation actually cross an `isic.3` boundary? A split whose
  children all reconverge into the same `isic.3` parent is imputed at
  the 4-digit level but **exact** at 3-digit and coarser (the `1/n`
  weights cancel out on re-aggregation); a split whose children land in
  *different* `isic.3` parents carries real, propagated allocation
  uncertainty. Only the second kind should really count as "consequential"
  imputation exposure. This is the same distinction the node-link
  diagrams in [Explorer](02-crossmap-explorer.qmd)/
  [Country variants](03-country-variants.qmd) show per component (a
  reconvergent vs. diverging fan of dashed split edges); this page shows
  its coverage-level rollup.

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
tile_any_split <- crossmap_collection$tile_any_split
tile_frac_split <- crossmap_collection$tile_frac_split
tile_frac_split_crossing <- crossmap_collection$tile_frac_split_crossing
ctr_codes <- crossmap_collection$ctr_codes
```

All three tiles below share one country ordering -- by each country's
*overall* (all-years-pooled) fraction of `isic` values in a crossing
component, weighted by volume (`sum(n_crossing) / sum(n_isic)`, not a
plain per-year average, so a country's one thin reporting year can't
skew its rank) -- rather than each plot re-sorting on its own metric.
That keeps a given row meaning "this country" consistently across all
three figures, so you can read the same row down the page and see
stage-1 coverage vs. stage-1+2 consequential coverage for the same
country side by side.

```{r}
country_order <- tile_frac_split_crossing |>
  group_by(country) |>
  summarise(total_isic = sum(n_isic), total_crossing = sum(n_crossing), .groups = "drop") |>
  mutate(agg_frac_crossing = total_crossing / total_isic) |>
  arrange(agg_frac_crossing) |>
  pull(country)
```

## Stage 1: Split Coverage (isiccomb -> isic)

Two views of the same underlying fact, at different resolutions: the
binary tile is a strict coarsening of the continuous one
(`any_split == (frac_split > 0)`) -- read them as a pair, binary for
"where does this even come up," continuous for "how much does it matter
there."

### Any Split?

```{r}
tile_any_split |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  mutate(country = factor(country, levels = country_order)) |>
  ggplot(aes(x = as.factor(year), y = country, fill = any_split)) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_manual(values = c("TRUE" = "#ef8a62", "FALSE" = "#67a9cf")) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_blank()) +
  labs(
    x = "Year", y = "Country", fill = "Any split?",
    title = "Countries reporting isiccomb combination codes",
    subtitle = "INDSTAT 4 REV. 3 Output data -- stage 1 only"
  )
```

### Fraction Split

```{r}
tile_frac_split |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  mutate(country = factor(country, levels = country_order)) |>
  ggplot(aes(x = as.factor(year), y = country, fill = frac_split)) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_blank()) +
  labs(
    x = "Year", y = "Country", fill = "Fraction split",
    title = "Share of isic values derived from an isiccomb split",
    subtitle = "INDSTAT 4 REV. 3 Output data -- stage 1 only"
  )
```

## Stage 1+2: Crossing isic.3

`frac_split_crossing` shares `frac_split`'s denominator (fraction of a
country-year's `isic` values) but restricts the numerator to values
whose combo group *crosses* an `isic.3` boundary -- i.e. it's the
"consequential" subset of stage 1's coverage, once the second stage is
accounted for: `frac_split = frac_split_reconverging + frac_split_crossing`.

```{r}
tile_frac_split_crossing |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  mutate(country = factor(country, levels = country_order)) |>
  ggplot(aes(x = as.factor(year), y = country, fill = frac_split_crossing)) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_distiller(palette = "Oranges", direction = 1) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_blank()) +
  labs(
    x = "Year", y = "Country", fill = "Fraction crossing",
    title = "Share of isic values with real, propagated allocation uncertainty",
    subtitle = "INDSTAT 4 REV. 3 Output data -- stage 1+2 composed"
  )
```

How much of stage 1's coverage is actually consequential, in aggregate
across the whole dataset:

```{r}
overall <- tile_frac_split |>
  left_join(tile_frac_split_crossing |> select(country, year, frac_split_crossing),
    by = c("country", "year")
  ) |>
  summarise(
    total_isic = sum(n_isic),
    total_split = sum(n_split),
    total_crossing = sum(round(frac_split_crossing * n_isic)),
  )

overall |>
  mutate(
    pct_of_all_isic_split = 100 * total_split / total_isic,
    pct_of_all_isic_crossing = 100 * total_crossing / total_isic,
    pct_of_split_that_crosses = 100 * total_crossing / total_split
  )
```

## Biggest Overstatements

Aggregating (all years pooled, volume-weighted -- same approach as
`country_order` above) down to one row per country makes it possible to
ask directly: for which reporters is stage-1 `frac_split` mostly noise
(reconvergent diamonds), and for which is it mostly signal (genuine
crossing)? Restricted to countries with at least 5 split rows pooled
across all their years, so a single small sample can't produce a
misleadingly clean 0% or 100%.

```{r}
country_split_summary <- tile_frac_split |>
  left_join(tile_frac_split_crossing |> select(country, year, n_crossing),
    by = c("country", "year")
  ) |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  group_by(country, name) |>
  summarise(
    n_years = n(),
    total_split = sum(n_split),
    total_crossing = sum(n_crossing),
    .groups = "drop"
  ) |>
  mutate(
    total_reconverging = total_split - total_crossing,
    pct_split_reconverging = 100 * total_reconverging / total_split
  ) |>
  filter(total_split >= 5)
```

Ranked by **absolute overstatement**: how much stage-1 coverage this
country would report vs. how much of that is actually consequential
(largest gap first -- i.e. the countries where looking only at the
first tile above would mislead you the most):

```{r}
country_split_summary |>
  left_join(
    tile_frac_split |> group_by(country) |> summarise(total_isic = sum(n_isic), .groups = "drop"),
    by = "country"
  ) |>
  mutate(
    frac_split = total_split / total_isic,
    frac_crossing = total_crossing / total_isic,
    gap = frac_split - frac_crossing
  ) |>
  arrange(desc(gap)) |>
  transmute(
    country, name, n_years, total_split, total_reconverging,
    pct_split_reconverging = round(pct_split_reconverging, 1),
    frac_split = round(frac_split, 3),
    frac_crossing = round(frac_crossing, 3),
    gap = round(gap, 3)
  ) |>
  head(15) |>
  knitr::kable()
```

Ranked by **share of splitting that's reconvergent**, i.e. "lots of
reconvergent diamonds" directly -- countries at 100% have *zero* real
crossing exposure despite a non-trivial `frac_split`:

```{r}
country_split_summary |>
  arrange(desc(pct_split_reconverging)) |>
  transmute(country, name, n_years, total_split, total_reconverging,
    pct_split_reconverging = round(pct_split_reconverging, 1)
  ) |>
  head(15) |>
  knitr::kable()
```
