---
title: "crossmap explorer: isiccomb splits + isic.3 aggregation, per country-year"
execute:
  warning: false
  cache: true
---

## Motivation

Component (c): zoom into a single country-year's crossmap collection.
(Component (d), the whole-classification overview, is a separate page --
see [Composed overview](04-composed-overview.qmd), built on
`xmap::compose_xmap()`,
[xmap#29](https://github.com/cynthiahqy/xmap/issues/29).)

Node-link crossmap design from Huang (2023),
[arxiv:2308.06535](https://arxiv.org/html/2308.06535v1) -- the design
behind the `xmap` package itself -- rather than a Sankey diagram:

- **source node text**: *italic* for one-to-many (splitting) sources,
  **bold** for one-to-one sources
- **edge line style**: dashed for one-to-many splits, solid for
  one-to-one relations, instead of Sankey's ribbon-width-as-weight
- **edge weight labels**: shown as text on the edge; dropped on solid
  (one-to-one) edges to reduce clutter, since that weight just repeats
  the incoming split edge's label
- **target node shading**: opacity/color scaled by in-degree (# of
  incoming edges), to show how "synthetic" a target category is

This is a **3-column chained node-link diagram** per `isiccomb` code:
`isiccomb -> isic member -> isic.3 parent`, combining two crossmaps (the
split, and the 3-vs-4-digit aggregation) into one figure, since the
middle column (`isic`) is simultaneously the target of the first
crossmap and the source of the second.

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
source(here("R", "prep.R")) # collapse_years()
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
combo_edges <- crossmap_collection$combo_edges
ctr_year_signature <- crossmap_collection$ctr_year_signature
max_isic3_indegree <- crossmap_collection$max_isic3_indegree
```

## Prep data for the diagram

Many country-years share the exact same `isiccomb` crossmap structure
(e.g. the same reporter using the same national classification variant
for a run of consecutive years). Rendering one diagram per country-year
would repeat identical figures dozens of times, so we first collapse
country-years into groups that share an *identical* crossmap (same set
of `isiccomb -> isic` edges with the same weights) and show each
distinct pattern only once, labelled with which country-years share it.

This groups *across* countries -- see
[Country variants](03-country-variants.qmd) instead for "all the variants
a single country used over time" (e.g. all the crossmaps DEU has had).

```{r}
## group country-years sharing an identical signature; one representative
## (country, year) per group is enough to render the diagram, since every
## member of the group has the same edges by construction
crossmap_group_list <- ctr_year_signature |>
  group_by(signature) |>
  group_modify(~ {
    by_ctr <- .x |>
      group_by(country, name) |>
      summarise(year_range = collapse_years(year), n_yr = n(), .groups = "drop") |>
      arrange(desc(n_yr))
    tibble(
      country = .x$country[1],
      year = .x$year[1],
      n_combo = .x$n_combo[1],
      n_crossing = .x$n_crossing[1],
      n_ctryr = nrow(.x),
      detail = paste(paste0(by_ctr$name, " (", by_ctr$country, "): ", by_ctr$year_range), collapse = "; ")
    )
  }) |>
  ungroup() |>
  arrange(desc(n_crossing), desc(n_combo), desc(n_ctryr)) |>
  mutate(label = paste0(
    "[", n_combo, " isiccomb, ", n_crossing, " crossing, ", n_ctryr, " country-yrs] ", detail
  ))

crossmap_group_list |> select(-signature)
```

```{r}
#| label: send-to-ojs
#| cache: false
ojs_define(crossmap_group_list = crossmap_group_list)
ojs_define(combo_edges = combo_edges)
ojs_define(max_isic3_indegree = max_isic3_indegree)
```

## Node-link diagram

```{ojs}
//| label: nodelink-inputs
d3 = require("d3@7")
import { buildCrossmapNodeLink3Col } from "./js/node-link.js"

viewof selectedCtrYr = Inputs.select(
  new Map(transpose(crossmap_group_list).map(d => [d.label, { country: d.country, year: d.year }])),
  { sort: false }
)
```

```{ojs}
//| label: nodelink-diagram
filteredEdges = transpose(combo_edges).filter(
  d => d.country === selectedCtrYr.country && d.year === selectedCtrYr.year
)

comboGroups = d3.sort(
  d3.groups(filteredEdges, d => d.isiccomb),
  ([, rows]) => -(new Set(rows.map(r => r.isic3)).size)
)

nodeLinkSmallMultiples = html`<div style="display:flex; flex-wrap:wrap; gap:1rem; align-items:flex-start;">
  ${comboGroups.map(([isiccombCode, rows]) => {
    const nParents = new Set(rows.map(r => r.isic3)).size;
    const status = nParents > 1
      ? `crosses ${nParents} isic.3 groups`
      : "single isic.3 group";
    return html`<div style="border:1px solid #ddd; border-radius:6px; padding:0.5rem;">
      <div style="font: 600 12px sans-serif; margin-bottom:4px;">${isiccombCode} (${status})</div>
      ${buildCrossmapNodeLink3Col(d3, rows, { maxIsic3Indegree: max_isic3_indegree })}
    </div>`;
  })}
</div>`

nodeLinkSmallMultiples
```
