---
title: "crossmap explorer: isiccomb splits + isic.3 aggregation, per country-year"
execute:
  warning: false
  cache: true
---

## Motivation

Component (c): zoom into a single country-year's crossmap collection.
(Component (d), the whole-classification overview, is a separate page --
see `index.qmd` -- blocked on `xmap::compose_xmap()`,
[xmap#29](https://github.com/cynthiahqy/xmap/issues/29).)

Node-link crossmap design from Huang (2023),
[arxiv:2308.06535](https://arxiv.org/html/2308.06535v1) -- the design
behind the `xmap` package itself -- rather than a Sankey diagram:

- **source node text**: *italic* for one-to-many (splitting) sources,
  **bold** for one-to-one sources
- **edge line style**: dashed for one-to-many splits, solid for
  one-to-one relations, instead of Sankey's ribbon-width-as-weight
- **edge weight labels**: shown as text on the edge; dropped on solid
  (one-to-one) edges to reduce clutter, since that weight just repeats
  the incoming split edge's label
- **target node shading**: opacity/color scaled by in-degree (# of
  incoming edges), to show how "synthetic" a target category is

This is a **3-column chained node-link diagram** per `isiccomb` code:
`isiccomb -> isic member -> isic.3 parent`, combining two crossmaps (the
split, and the 3-vs-4-digit aggregation) into one figure, since the
middle column (`isic`) is simultaneously the target of the first
crossmap and the source of the second.

## Setup

```{r}
library(tidyverse)
library(here)
library(targets)
source(here("R", "prep.R")) # collapse_years()
```

```{r}
combo_edges <- tar_read(combo_edges)
ctr_year_signature <- tar_read(ctr_year_signature)
```

## Prep data for the diagram

Many country-years share the exact same `isiccomb` crossmap structure
(e.g. the same reporter using the same national classification variant
for a run of consecutive years). Rendering one diagram per country-year
would repeat identical figures dozens of times, so we first collapse
country-years into groups that share an *identical* crossmap (same set
of `isiccomb -> isic` edges with the same weights) and show each
distinct pattern only once, labelled with which country-years share it.

This groups *across* countries -- see
[Country variants](03-country-variants.qmd) instead for "all the variants
a single country used over time" (e.g. all the crossmaps DEU has had).

```{r}
## group country-years sharing an identical signature; one representative
## (country, year) per group is enough to render the diagram, since every
## member of the group has the same edges by construction
crossmap_group_list <- ctr_year_signature |>
  group_by(signature) |>
  group_modify(~ {
    by_ctr <- .x |>
      group_by(country, name) |>
      summarise(year_range = collapse_years(year), n_yr = n(), .groups = "drop") |>
      arrange(desc(n_yr))
    tibble(
      country = .x$country[1],
      year = .x$year[1],
      n_combo = .x$n_combo[1],
      n_crossing = .x$n_crossing[1],
      n_ctryr = nrow(.x),
      detail = paste(paste0(by_ctr$name, " (", by_ctr$country, "): ", by_ctr$year_range), collapse = "; ")
    )
  }) |>
  ungroup() |>
  arrange(desc(n_crossing), desc(n_combo), desc(n_ctryr)) |>
  mutate(label = paste0(
    "[", n_combo, " isiccomb, ", n_crossing, " crossing, ", n_ctryr, " country-yrs] ", detail
  ))

crossmap_group_list |> select(-signature)
```

```{r}
#| label: send-to-ojs
#| cache: false
ojs_define(crossmap_group_list = crossmap_group_list)
ojs_define(combo_edges = combo_edges)
```

## Node-link diagram

```{ojs}
//| label: nodelink-inputs
d3 = require("d3@7")

viewof selectedCtrYr = Inputs.select(
  new Map(transpose(crossmap_group_list).map(d => [d.label, { country: d.country, year: d.year }])),
  { sort: false }
)
```

```{ojs}
//| label: nodelink-diagram
filteredEdges = transpose(combo_edges).filter(
  d => d.country === selectedCtrYr.country && d.year === selectedCtrYr.year
)

comboGroups = d3.sort(
  d3.groups(filteredEdges, d => d.isiccomb),
  ([, rows]) => -(new Set(rows.map(r => r.isic3)).size)
)

function evenY(n, height, pad = 18) {
  if (n <= 1) return [height / 2];
  const usable = height - 2 * pad;
  return d3.range(n).map(i => pad + (i * usable) / (n - 1));
}

function buildCrossmapNodeLink(edgesForCode) {
  const isiccomb = edgesForCode[0].isiccomb;

  // col0: the combo code (always one-to-many here, since it's a crossing code)
  const col0 = [isiccomb];

  // col1: isic members, grouped by their isic.3 parent to reduce edge crossings
  const col1 = Array.from(new Set(edgesForCode.map(d => d.isic))).sort((a, b) => {
    const pa = edgesForCode.find(d => d.isic === a).isic3;
    const pb = edgesForCode.find(d => d.isic === b).isic3;
    return d3.ascending(pa, pb) || d3.ascending(a, b);
  });

  // col2: isic.3 parents, most-children (most "synthetic") first
  const inDeg2 = d3.rollup(edgesForCode, v => new Set(v.map(d => d.isic)).size, d => d.isic3);
  const col2 = Array.from(new Set(edgesForCode.map(d => d.isic3)))
    .sort((a, b) => d3.descending(inDeg2.get(a), inDeg2.get(b)) || d3.ascending(a, b));

  const width = 380;
  const nRows = Math.max(col1.length, col2.length);
  const height = Math.max(90, nRows * 20 + 36);

  const x0 = 54, x1 = width / 2, x2 = width - 54;
  const y0 = evenY(1, height);
  const y1 = evenY(col1.length, height);
  const y2 = evenY(col2.length, height);

  const pos0 = new Map(col0.map((d, i) => [d, y0[i]]));
  const pos1 = new Map(col1.map((d, i) => [d, y1[i]]));
  const pos2 = new Map(col2.map((d, i) => [d, y2[i]]));

  const maxIn2 = d3.max(col2, d => inDeg2.get(d)) ?? 1;
  const shade = d3.scaleSequential(d3.interpolateBlues).domain([0, maxIn2]);

  const svg = d3.create("svg")
    .attr("viewBox", [0, 0, width, height])
    .attr("width", width)
    .attr("height", height)
    .attr("style", "max-width: 100%; height: auto; font: 10px sans-serif;");

  // stage 1 edges: isiccomb -> isic (dashed, one-to-many split, weight labelled)
  svg.append("g")
    .selectAll("path")
    .data(edgesForCode)
    .join("path")
      .attr("d", d => {
        const sy = pos0.get(d.isiccomb), ty = pos1.get(d.isic);
        return `M${x0},${sy} C${(x0 + x1) / 2},${sy} ${(x0 + x1) / 2},${ty} ${x1},${ty}`;
      })
      .attr("fill", "none")
      .attr("stroke", "#888")
      .attr("stroke-width", 1)
      .attr("stroke-dasharray", "3,2")
    .append("title")
      .text(d => `${d.isiccomb} -> ${d.isic}\nweight: ${d.weight.toFixed(3)}`);

  // stage 1 weight labels
  svg.append("g")
    .selectAll("text")
    .data(edgesForCode)
    .join("text")
      .attr("x", (x0 + x1) / 2)
      .attr("y", d => (pos0.get(d.isiccomb) + pos1.get(d.isic)) / 2 - 2)
      .attr("text-anchor", "middle")
      .attr("fill", "#666")
      .text(d => d.weight.toFixed(2));

  // stage 2 edges: isic -> isic.3 (solid, one-to-one, no label -- repeats stage 1 weight)
  const stage2 = Array.from(new Map(edgesForCode.map(d => [d.isic, d])).values());
  svg.append("g")
    .selectAll("path")
    .data(stage2)
    .join("path")
      .attr("d", d => {
        const sy = pos1.get(d.isic), ty = pos2.get(d.isic3);
        return `M${x1},${sy} C${(x1 + x2) / 2},${sy} ${(x1 + x2) / 2},${ty} ${x2},${ty}`;
      })
      .attr("fill", "none")
      .attr("stroke", "#888")
      .attr("stroke-width", 1)
    .append("title")
      .text(d => `${d.isic} -> ${d.isic3}\nweight: ${d.weight.toFixed(3)}`);

  // col0 node: italic (one-to-many source)
  svg.append("circle").attr("cx", x0).attr("cy", y0[0]).attr("r", 3).attr("fill", "#333");
  svg.append("text")
    .attr("x", x0 - 6).attr("y", y0[0]).attr("dy", "0.32em")
    .attr("text-anchor", "end").attr("font-style", "italic").attr("font-weight", 600)
    .text(isiccomb);

  // col1 nodes: bold (one-to-one sources for stage 2)
  svg.append("g")
    .selectAll("circle")
    .data(col1)
    .join("circle")
      .attr("cx", x1).attr("cy", d => pos1.get(d)).attr("r", 2.5).attr("fill", "#333");
  svg.append("g")
    .selectAll("text")
    .data(col1)
    .join("text")
      .attr("x", x1).attr("y", d => pos1.get(d)).attr("dy", "-0.5em")
      .attr("text-anchor", "middle").attr("font-weight", 600)
      .text(d => d);

  // col2 nodes: shaded by in-degree ("how synthetic")
  svg.append("g")
    .selectAll("circle")
    .data(col2)
    .join("circle")
      .attr("cx", x2).attr("cy", d => pos2.get(d)).attr("r", 6)
      .attr("fill", d => shade(inDeg2.get(d)))
      .attr("stroke", "#333").attr("stroke-width", 0.5)
    .append("title")
      .text(d => `${d} -- ${inDeg2.get(d)} isic member(s) feed into this parent`);
  svg.append("g")
    .selectAll("text")
    .data(col2)
    .join("text")
      .attr("x", x2 + 10).attr("y", d => pos2.get(d)).attr("dy", "0.32em")
      .attr("text-anchor", "start")
      .text(d => d);

  return svg.node();
}

nodeLinkSmallMultiples = html`<div style="display:flex; flex-wrap:wrap; gap:1rem; align-items:flex-start;">
  ${comboGroups.map(([isiccombCode, rows]) => {
    const nParents = new Set(rows.map(r => r.isic3)).size;
    const status = nParents > 1
      ? `crosses ${nParents} isic.3 groups`
      : "single isic.3 group";
    return html`<div style="border:1px solid #ddd; border-radius:6px; padding:0.5rem;">
      <div style="font: 600 12px sans-serif; margin-bottom:4px;">${isiccombCode} (${status})</div>
      ${buildCrossmapNodeLink(rows)}
    </div>`;
  })}
</div>`

nodeLinkSmallMultiples
```
