---
title: "country variants: all the crossmaps one country has used, over time"
execute:
  warning: false
  cache: true
---

## Motivation

[Explorer](02-crossmap-explorer.qmd) picks any single country-year from
the globally-deduplicated list of 218 distinct crossmap patterns. This
page asks a different question: **for one country, how many genuinely
different `isiccomb -> isic` crossmaps did it use across all its
reporting years, and when did each apply?**

A country's classification/reporting scheme is not always fixed over
time -- e.g. DEU (Germany, code `276`) turns out to have used 3
structurally distinct crossmaps across its 13 reporting years, not 13
independent ones. This groups a single country's years by identical
crossmap signature -- the same de-duplication as the Explorer page, just
sliced by country first instead of pooled across all countries.

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
source(here("R", "prep.R")) # collapse_years()
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
combo_edges <- crossmap_collection$combo_edges
max_isic3_indegree <- crossmap_collection$max_isic3_indegree
ctr_year_signature <- crossmap_collection$ctr_year_signature
```

## Prep data

```{r}
## countries with at least one isiccomb-reporting year
country_list <- ctr_year_signature |>
  distinct(country, name) |>
  arrange(name)

country_list
```

```{r}
## within each country, collapse years sharing an identical crossmap
## signature into one "variant" -- year_range covers all years using
## that variant, first_year is the representative used to pull edges
## for the diagram (every year in the group has the same edges by
## construction, so any one of them will do)
country_variants <- ctr_year_signature |>
  group_by(country, name, signature) |>
  summarise(
    year_range = collapse_years(year),
    n_yr = n(),
    first_year = min(year),
    n_combo = n_combo[1],
    n_crossing = n_crossing[1],
    .groups = "drop"
  ) |>
  arrange(country, first_year) |>
  mutate(variant_label = paste0(
    year_range, "  [", n_combo, " isiccomb, ", n_crossing, " crossing, ", n_yr, " yr]"
  ))

country_variants |> select(-signature)
```

```{r}
#| label: send-to-ojs
#| cache: false
ojs_define(country_list = country_list)
ojs_define(country_variants = country_variants)
ojs_define(combo_edges = combo_edges)
ojs_define(max_isic3_indegree = max_isic3_indegree)
```

## Variants over time

```{ojs}
//| label: country-inputs
d3 = require("d3@7")
import { buildCrossmapNodeLink3Col } from "./js/node-link.js"

viewof selectedCountry = Inputs.select(
  new Map(transpose(country_list).map(d => [`${d.name} (${d.country})`, d.country])),
  { sort: false }
)
```

```{ojs}
//| label: country-diagram
selectedVariants = transpose(country_variants).filter(d => d.country === selectedCountry)

function variantSection(variant) {
  const edgesForVariant = transpose(combo_edges).filter(
    d => d.country === variant.country && d.year === variant.first_year
  );
  const comboGroups = d3.sort(
    d3.groups(edgesForVariant, d => d.isiccomb),
    ([, rows]) => -(new Set(rows.map(r => r.isic3)).size)
  );

  return html`<section style="margin-bottom:1.5rem;">
    <h3 style="font-size:14px; margin-bottom:0.5rem;">${variant.variant_label}</h3>
    <div style="display:flex; flex-wrap:wrap; gap:0.75rem; align-items:flex-start;">
      ${comboGroups.map(([isiccombCode, rows]) => {
        const nParents = new Set(rows.map(r => r.isic3)).size;
        const status = nParents > 1 ? `crosses ${nParents} isic.3 groups` : "single isic.3 group";
        return html`<div style="border:1px solid #ddd; border-radius:6px; padding:0.4rem;">
          <div style="font: 600 11px sans-serif; margin-bottom:3px;">${isiccombCode} (${status})</div>
          ${buildCrossmapNodeLink3Col(d3, rows, {
            maxIsic3Indegree: max_isic3_indegree,
            width: 340, fontSize: 9, rowHeight: 18, padY: 32, minHeight: 80, margin: 50,
          })}
        </div>`;
      })}
    </div>
  </section>`;
}

countryVariantsView = html`<div>
  ${selectedVariants.map(v => variantSection(v))}
</div>`

countryVariantsView
```
