---
title: "country variants: all the crossmaps one country has used, over time"
execute:
  warning: false
  cache: true
---

## Motivation

[Explorer](02-crossmap-explorer.qmd) picks any single country-year from
the globally-deduplicated list of 218 distinct crossmap patterns. This
page asks a different question: **for one country, how many genuinely
different `isiccomb -> isic` crossmaps did it use across all its
reporting years, and when did each apply?**

A country's classification/reporting scheme is not always fixed over
time -- e.g. DEU (Germany, code `276`) turns out to have used 3
structurally distinct crossmaps across its 13 reporting years, not 13
independent ones. This groups a single country's years by identical
crossmap signature -- the same de-duplication as the Explorer page, just
sliced by country first instead of pooled across all countries.

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
source(here("R", "prep.R")) # collapse_years()
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
combo_edges <- crossmap_collection$combo_edges
max_isic3_indegree <- crossmap_collection$max_isic3_indegree
ctr_year_signature <- crossmap_collection$ctr_year_signature
```

## Prep data

```{r}
## countries with at least one isiccomb-reporting year
country_list <- ctr_year_signature |>
  distinct(country, name) |>
  arrange(name)

country_list
```

```{r}
## within each country, collapse years sharing an identical crossmap
## signature into one "variant" -- year_range covers all years using
## that variant, first_year is the representative used to pull edges
## for the diagram (every year in the group has the same edges by
## construction, so any one of them will do)
country_variants <- ctr_year_signature |>
  group_by(country, name, signature) |>
  summarise(
    year_range = collapse_years(year),
    n_yr = n(),
    first_year = min(year),
    n_combo = n_combo[1],
    n_crossing = n_crossing[1],
    .groups = "drop"
  ) |>
  arrange(country, first_year) |>
  mutate(variant_label = paste0(
    year_range, "  [", n_combo, " isiccomb, ", n_crossing, " crossing, ", n_yr, " yr]"
  ))

country_variants |> select(-signature)
```

```{r}
#| label: send-to-ojs
#| cache: false
ojs_define(country_list = country_list)
ojs_define(country_variants = country_variants)
ojs_define(combo_edges = combo_edges)
ojs_define(max_isic3_indegree = max_isic3_indegree)
```

## Variants over time

```{ojs}
//| label: country-inputs
d3 = require("d3@7")

viewof selectedCountry = Inputs.select(
  new Map(transpose(country_list).map(d => [`${d.name} (${d.country})`, d.country])),
  { sort: false }
)
```

```{ojs}
//| label: country-diagram
selectedVariants = transpose(country_variants).filter(d => d.country === selectedCountry)

function evenY(n, height, pad = 18) {
  if (n <= 1) return [height / 2];
  const usable = height - 2 * pad;
  return d3.range(n).map(i => pad + (i * usable) / (n - 1));
}

function buildCrossmapNodeLink(edgesForCode) {
  const isiccomb = edgesForCode[0].isiccomb;

  // col0: the combo code
  const col0 = [isiccomb];

  // col1: isic members, grouped by their isic.3 parent to reduce edge crossings
  const col1 = Array.from(new Set(edgesForCode.map(d => d.isic))).sort((a, b) => {
    const pa = edgesForCode.find(d => d.isic === a).isic3;
    const pb = edgesForCode.find(d => d.isic === b).isic3;
    return d3.ascending(pa, pb) || d3.ascending(a, b);
  });

  // col2: isic.3 parents, most-children (most "synthetic") first
  const inDeg2 = d3.rollup(edgesForCode, v => new Set(v.map(d => d.isic)).size, d => d.isic3);
  const col2 = Array.from(new Set(edgesForCode.map(d => d.isic3)))
    .sort((a, b) => d3.descending(inDeg2.get(a), inDeg2.get(b)) || d3.ascending(a, b));

  const width = 340;
  const nRows = Math.max(col1.length, col2.length);
  const height = Math.max(80, nRows * 18 + 32);

  const x0 = 50, x1 = width / 2, x2 = width - 50;
  const y0 = evenY(1, height);
  const y1 = evenY(col1.length, height);
  const y2 = evenY(col2.length, height);

  const pos0 = new Map(col0.map((d, i) => [d, y0[i]]));
  const pos1 = new Map(col1.map((d, i) => [d, y1[i]]));
  const pos2 = new Map(col2.map((d, i) => [d, y2[i]]));

  // fixed domain (not renormalized per component) -- see
  // 02-crossmap-explorer.qmd for why: a plain 1-to-1 target in a
  // component with no aggregation must not render as dark as a genuinely
  // aggregated target elsewhere just because it's the local max.
  const shade = d3.scaleSequential(d3.interpolateBlues).domain([0, max_isic3_indegree]);

  const svg = d3.create("svg")
    .attr("viewBox", [0, 0, width, height])
    .attr("width", width)
    .attr("height", height)
    .attr("style", "max-width: 100%; height: auto; font: 9px sans-serif;");

  // stage 1: isiccomb -> isic (dashed, split, weight labelled)
  svg.append("g")
    .selectAll("path")
    .data(edgesForCode)
    .join("path")
      .attr("d", d => {
        const sy = pos0.get(d.isiccomb), ty = pos1.get(d.isic);
        return `M${x0},${sy} C${(x0 + x1) / 2},${sy} ${(x0 + x1) / 2},${ty} ${x1},${ty}`;
      })
      .attr("fill", "none").attr("stroke", "#888").attr("stroke-width", 1)
      .attr("stroke-dasharray", "3,2")
    .append("title")
      .text(d => `${d.isiccomb} -> ${d.isic}\nweight: ${d.weight.toFixed(3)}`);

  svg.append("g")
    .selectAll("text")
    .data(edgesForCode)
    .join("text")
      .attr("x", (x0 + x1) / 2)
      .attr("y", d => (pos0.get(d.isiccomb) + pos1.get(d.isic)) / 2 - 2)
      .attr("text-anchor", "middle").attr("fill", "#666")
      .text(d => d.weight.toFixed(2));

  // stage 2: isic -> isic.3 (solid, one-to-one, no label)
  const stage2 = Array.from(new Map(edgesForCode.map(d => [d.isic, d])).values());
  svg.append("g")
    .selectAll("path")
    .data(stage2)
    .join("path")
      .attr("d", d => {
        const sy = pos1.get(d.isic), ty = pos2.get(d.isic3);
        return `M${x1},${sy} C${(x1 + x2) / 2},${sy} ${(x1 + x2) / 2},${ty} ${x2},${ty}`;
      })
      .attr("fill", "none").attr("stroke", "#888").attr("stroke-width", 1)
    .append("title")
      .text(d => `${d.isic} -> ${d.isic3}\nweight: ${d.weight.toFixed(3)}`);

  // col0: italic (one-to-many source)
  svg.append("circle").attr("cx", x0).attr("cy", y0[0]).attr("r", 3).attr("fill", "#333");
  svg.append("text")
    .attr("x", x0 - 6).attr("y", y0[0]).attr("dy", "0.32em")
    .attr("text-anchor", "end").attr("font-style", "italic").attr("font-weight", 600)
    .text(isiccomb);

  // col1: bold (one-to-one sources for stage 2)
  svg.append("g").selectAll("circle").data(col1).join("circle")
    .attr("cx", x1).attr("cy", d => pos1.get(d)).attr("r", 2.5).attr("fill", "#333");
  svg.append("g").selectAll("text").data(col1).join("text")
    .attr("x", x1).attr("y", d => pos1.get(d)).attr("dy", "-0.5em")
    .attr("text-anchor", "middle").attr("font-weight", 600)
    .text(d => d);

  // col2: shaded by in-degree ("how synthetic")
  svg.append("g").selectAll("circle").data(col2).join("circle")
    .attr("cx", x2).attr("cy", d => pos2.get(d)).attr("r", 6)
    .attr("fill", d => shade(inDeg2.get(d)))
    .attr("stroke", "#333").attr("stroke-width", 0.5)
    .append("title")
      .text(d => `${d} -- ${inDeg2.get(d)} isic member(s) feed into this parent`);
  svg.append("g").selectAll("text").data(col2).join("text")
    .attr("x", x2 + 10).attr("y", d => pos2.get(d)).attr("dy", "0.32em")
    .attr("text-anchor", "start")
    .text(d => d);

  return svg.node();
}

function variantSection(variant) {
  const edgesForVariant = transpose(combo_edges).filter(
    d => d.country === variant.country && d.year === variant.first_year
  );
  const comboGroups = d3.sort(
    d3.groups(edgesForVariant, d => d.isiccomb),
    ([, rows]) => -(new Set(rows.map(r => r.isic3)).size)
  );

  return html`<section style="margin-bottom:1.5rem;">
    <h3 style="font-size:14px; margin-bottom:0.5rem;">${variant.variant_label}</h3>
    <div style="display:flex; flex-wrap:wrap; gap:0.75rem; align-items:flex-start;">
      ${comboGroups.map(([isiccombCode, rows]) => {
        const nParents = new Set(rows.map(r => r.isic3)).size;
        const status = nParents > 1 ? `crosses ${nParents} isic.3 groups` : "single isic.3 group";
        return html`<div style="border:1px solid #ddd; border-radius:6px; padding:0.4rem;">
          <div style="font: 600 11px sans-serif; margin-bottom:3px;">${isiccombCode} (${status})</div>
          ${buildCrossmapNodeLink(rows)}
        </div>`;
      })}
    </div>
  </section>`;
}

countryVariantsView = html`<div>
  ${selectedVariants.map(v => variantSection(v))}
</div>`

countryVariantsView
```
