---
title: "Composed Overview"
execute:
  warning: false
  cache: true
---

## Motivation

Component (d): the composed, whole-classification `isiccomb -> isic.3`
view -- one graph, not per country-year, showing every place in the
classification where a real one-to-many split or many-to-one aggregation
happens, with the pass-through-unchanged majority collapsed out of view.

Composed with [`xmap::compose_xmap()`](https://github.com/cynthiahqy/xmap/issues/29)
in [Extraction](00-extract-crossmaps.qmd), chaining `isiccomb -> isic` and
`isic -> isic.3` through their shared `isic` intermediate into a single
`isiccomb -> isic.3` crossmap -- without materialising `isic`-level values,
which aren't meaningful on their own (they're an artifact of the naive
split). This is the same information [Explorer](02-crossmap-explorer.qmd)/
[Country variants](03-country-variants.qmd) show per country-year
component, at the whole-classification level: not "what happened in one
country-year" but "which `isiccomb` codes, anywhere in the dataset,
actually carry a real split or aggregation."

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
source(here("R", "prep.R")) # collapse_years()
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
composed_isiccomb_isic3 <- crossmap_collection$composed_isiccomb_isic3
ctr_codes <- crossmap_collection$ctr_codes
```

## Preparing the Diagram

Every `isiccomb` code turns out to compose the exact same way in every
country-year that reports it (see [Extraction](00-extract-crossmaps.qmd)),
so the composed crossmap collapses to one global structure with no extra
reconciliation -- distinct `(isiccomb, isic3, weight)` triples, bundling
away the pass-through-unchanged majority (a code mapping to exactly one
`isic.3` parent at weight 1) and showing only codes that cross an
`isic.3` boundary.

```{r}
## dedup: every isiccomb code composes identically in every country-year
## that reports it, so a global distinct() is a lossless reduction
composed_code_edges <- composed_isiccomb_isic3 |>
  distinct(isiccomb, isic3, weight) |>
  add_count(isiccomb, name = "n_isic3")

n_codes <- n_distinct(composed_code_edges$isiccomb)
n_bundled <- composed_code_edges |> filter(n_isic3 == 1) |> nrow()
n_crossing <- n_codes - n_bundled

composed_crossing_edges <- composed_code_edges |>
  filter(n_isic3 > 1) |>
  arrange(desc(n_isic3), isiccomb)

glue::glue(
  "{n_codes} distinct isiccomb codes in the classification: ",
  "{n_bundled} map one-to-one onto a single isic.3 parent (bundled, not shown below); ",
  "{n_crossing} cross into more than one isic.3 parent."
)
```

Deduping away `country`/`year` above is lossless for the *structure* of each
component, but it also erases which slice of the real dataset each small
multiple actually reflects -- a code reported by one country in one year
and a code reported by 50 country-years look identical otherwise. Same
country-year attribution as `crossmap_group_list`'s `detail` field in
[Explorer](02-crossmap-explorer.qmd), computed here per `isiccomb` instead
of per identical-crossmap group:

`coverage_summary` caps the always-visible subtitle at the top 3
countries by years reported (`+N more` beyond that) -- codes reported by
dozens of countries (e.g. `1810A`, 45 countries) would otherwise produce
a ~1300-character subtitle wrapping across a whole small multiple.
`coverage_detail` (full, untruncated) is kept for the hover tooltip.

```{r}
composed_code_coverage <- composed_isiccomb_isic3 |>
  distinct(country, year, isiccomb) |>
  semi_join(composed_crossing_edges, by = "isiccomb") |>
  left_join(ctr_codes, by = c("country" = "code")) |>
  group_by(isiccomb, country, name) |>
  summarise(year_range = collapse_years(year), n_yr = n(), .groups = "drop") |>
  arrange(isiccomb, desc(n_yr)) |>
  group_by(isiccomb) |>
  summarise(
    n_country = n_distinct(country),
    n_ctryr = sum(n_yr),
    coverage_detail = paste(paste0(name, " (", country, "): ", year_range), collapse = "; "),
    coverage_summary = {
      entries <- paste0(name, " (", country, "): ", year_range)
      shown <- entries[seq_len(min(3, length(entries)))]
      n_more <- length(entries) - length(shown)
      if (n_more > 0) paste0(paste(shown, collapse = "; "), "; +", n_more, " more")
      else paste(shown, collapse = "; ")
    },
    .groups = "drop"
  )

composed_code_coverage
```

```{r}
#| label: send-to-ojs
#| cache: false
ojs_define(composed_crossing_edges = composed_crossing_edges)
ojs_define(composed_code_coverage = composed_code_coverage)
```

## Node-Link Overview

```{ojs}
//| label: nodelink-inputs
d3 = require("d3@7")
import { buildComposedNodeLink2Col } from "./js/node-link.js"
```

```{ojs}
//| label: nodelink-diagram
edges = transpose(composed_crossing_edges)
coverage = new Map(transpose(composed_code_coverage).map(d => [d.isiccomb, d]))

codeGroups = d3.sort(
  d3.groups(edges, d => d.isiccomb),
  ([, rows]) => -rows.length
)

nodeLinkOverview = html`<div style="display:flex; flex-wrap:wrap; gap:1rem; align-items:flex-start;">
  ${codeGroups.map(([isiccombCode, rows]) => {
    const cov = coverage.get(isiccombCode);
    const countLabel = `${cov.n_country} countr${cov.n_country === 1 ? "y" : "ies"}, ${cov.n_ctryr} country-yr${cov.n_ctryr === 1 ? "" : "s"}`;
    return html`<div style="border:1px solid #ddd; border-radius:6px; padding:0.5rem;" title="${countLabel}\n${cov.coverage_detail}">
      <div style="font: 600 12px sans-serif; margin-bottom:2px;">${isiccombCode} (${rows.length} isic.3 parents)</div>
      <div style="font: 10px sans-serif; color:#666; margin-bottom:4px;">${cov.coverage_summary}</div>
      ${buildComposedNodeLink2Col(d3, rows)}
    </div>`;
  })}
</div>`

nodeLinkOverview
```
