---
title: "composed overview: whole-classification isiccomb -> isic.3 structure"
execute:
  warning: false
  cache: true
---

## Motivation

Component (d): the composed, whole-classification `isiccomb -> isic.3`
view -- one graph, not per country-year, showing every place in the
classification where a real one-to-many split or many-to-one aggregation
happens, with the pass-through-unchanged majority collapsed out of view.

Composed with [`xmap::compose_xmap()`](https://github.com/cynthiahqy/xmap/issues/29)
in [Extraction](00-extract-crossmaps.qmd), chaining `isiccomb -> isic` and
`isic -> isic.3` through their shared `isic` intermediate into a single
`isiccomb -> isic.3` crossmap -- without materialising `isic`-level values,
which aren't meaningful on their own (they're an artifact of the naive
split). This is the same information [Explorer](02-crossmap-explorer.qmd)/
[Country variants](03-country-variants.qmd) show per country-year
component, at the whole-classification level: not "what happened in one
country-year" but "which `isiccomb` codes, anywhere in the dataset,
actually carry a real split or aggregation."

## Setup

Reads the crossmap collection prepared by
[Extraction](00-extract-crossmaps.qmd) -- render that page first if
`data/interim/crossmap_collection.rds` is missing or stale.

```{r}
library(tidyverse)
library(here)
```

```{r}
crossmap_collection <- readRDS(here("data", "interim", "crossmap_collection.rds"))
composed_isiccomb_isic3 <- crossmap_collection$composed_isiccomb_isic3
```

## Prep data for the diagram

Every `isiccomb` code turns out to compose the exact same way in every
country-year that reports it (see [Extraction](00-extract-crossmaps.qmd)),
so the composed crossmap collapses to one global structure with no extra
reconciliation -- distinct `(isiccomb, isic3, weight)` triples, bundling
away the pass-through-unchanged majority (a code mapping to exactly one
`isic.3` parent at weight 1) and showing only codes that cross an
`isic.3` boundary.

```{r}
## dedup: every isiccomb code composes identically in every country-year
## that reports it, so a global distinct() is a lossless reduction
composed_code_edges <- composed_isiccomb_isic3 |>
  distinct(isiccomb, isic3, weight) |>
  add_count(isiccomb, name = "n_isic3")

n_codes <- n_distinct(composed_code_edges$isiccomb)
n_bundled <- composed_code_edges |> filter(n_isic3 == 1) |> nrow()
n_crossing <- n_codes - n_bundled

composed_crossing_edges <- composed_code_edges |>
  filter(n_isic3 > 1) |>
  arrange(desc(n_isic3), isiccomb)

glue::glue(
  "{n_codes} distinct isiccomb codes in the classification: ",
  "{n_bundled} map one-to-one onto a single isic.3 parent (bundled, not shown below); ",
  "{n_crossing} cross into more than one isic.3 parent."
)
```

```{r}
#| label: send-to-ojs
#| cache: false
ojs_define(composed_crossing_edges = composed_crossing_edges)
```

## Node-link overview

```{ojs}
//| label: nodelink-inputs
d3 = require("d3@7")
import { buildComposedNodeLink2Col } from "./js/node-link.js"
```

```{ojs}
//| label: nodelink-diagram
edges = transpose(composed_crossing_edges)

codeGroups = d3.sort(
  d3.groups(edges, d => d.isiccomb),
  ([, rows]) => -rows.length
)

nodeLinkOverview = html`<div style="display:flex; flex-wrap:wrap; gap:1rem; align-items:flex-start;">
  ${codeGroups.map(([isiccombCode, rows]) => {
    return html`<div style="border:1px solid #ddd; border-radius:6px; padding:0.5rem;">
      <div style="font: 600 12px sans-serif; margin-bottom:4px;">${isiccombCode} (${rows.length} isic.3 parents)</div>
      ${buildComposedNodeLink2Col(d3, rows)}
    </div>`;
  })}
</div>`

nodeLinkOverview
```
